---
title: "COVID19 pandemic"
author: Kevin Leiss (https://twitter.com/kl_print)
date: 16.03.2020
output:
  html_document:
    df_print: paged
---
# Getting the data
```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning = FALSE, message = FALSE)
```

Data was collected from [https://github.com/CSSEGISandData/COVID-19](https://github.com/CSSEGISandData/COVID-19).

The code for this analysis is avaiable on [github](https://github.com/klprint/covid19).

```{r}
library(tidyverse)

cas = read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv") %>%
  gather("date", "cases", -c(`Province/State`,`Country/Region`, Lat, Long)) %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  group_by(`Country/Region`, date) %>%
  summarise(cases = sum(cases))

recov = read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Recovered.csv") %>%
  gather("date", "recovered", -c(`Province/State`,`Country/Region`, Lat, Long)) %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  group_by(`Country/Region`, date) %>%
  summarise(recovered = sum(recovered))

dea = read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Deaths.csv") %>%
  gather("date", "deaths", -c(`Province/State`,`Country/Region`, Lat, Long)) %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  group_by(`Country/Region`, date) %>%
  summarise(deaths = sum(deaths))


df =  dr = left_join(cas, recov) %>%
  left_join(dea) %>%
  mutate(death_rate = deaths/cases) %>%
  mutate(death_rate = ifelse(is.nan(death_rate), 0, death_rate)) %>%
  mutate(still_infected = cases - recovered - deaths) %>%
  group_by(`Country/Region`) %>%
  filter(max(cases) > 500)

df = df %>%
  select(-death_rate) %>%
  gather("group", "n", -c(`Country/Region`, date)) %>%
  as_tibble()
```


# Countries with more than 500 cases:
The following shows the onset of the COVID19 pandemic in all countries which reported up to today at least 500 cases, starting from March 01, 2020.

```{r, dpi=300, fig.width=10, fig.height=7}
per.country <- df %>%
  group_by(`Country/Region`) %>%
  filter(date > "2020-03-01") %>%
  ggplot(aes(x = date, y = n, color=group, group = group)) +
  geom_line(size=.25) +
  facet_wrap(vars(`Country/Region`), scales = "free_y") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5)) +
  ggtitle("COVID19 cases") +
  labs(subtitle="Counties w/ > 500 cases, 01 Mar. 2020 onward
Source: https://github.com/CSSEGISandData/COVID-19") +
  ylab("cases")

per.country
```

And on log-scale:

```{r , dpi=300, fig.width=10, fig.height=7}
per.country + scale_y_log10() + ylab("log10(cases)")
```

It is clearly visible, that in most of the countries the increase in cases is following an exponential growth.
Whoever wants to learn more about this, please check out Simon Anders nice [post](http://papagei.zmbh.uni-heidelberg.de/anders/covid/covid.html) and this [video](https://www.youtube.com/watch?v=Kas0tIxDvrg) of 3blue1brown.

# Prediction for Germany
If we assume an exponential growth, we can use the data available and see what to expect in the next couple of days.
I know that the circumstances can change very quickly, which is why I only provide the prediction for the next five days:

```{r}
x <- cas %>%
  filter(`Country/Region`=="Germany") %>%
  filter(date >= "2020-02-27") %>%
  mutate(d = as.numeric(date))

exp.fit <- nls(cases ~ A * exp( B * as.numeric( date
                                                - as.numeric(min(x$date) )
                                                , units = "days" )),
               start = c(A=20, B=.2),
               data = x)

x$exp.fit <- predict(exp.fit, x)

future <- tibble(
  date = min(x$date) + 0:(as.numeric(max(x$date) - min(x$date)) + 5)
) %>%
  mutate(cases = predict(exp.fit,.)) %>%
  add_column(type = "Prediction")

x <- rbind(
  x %>%
    ungroup() %>%
    select(date, cases) %>%
    add_column(type="Official"),
  future
)

x%>%
  ggplot(aes(x=date, y=cases, color=type)) +
  geom_point(data= function(y){
    y %>% filter(type=="Official")
  }, size=1) +
  geom_point(data= function(y){
    y %>% filter(type=="Prediction")
  }, size=.75)+
  ggtitle("Cases in Germany") +
  xlab("Date") +
  ylab("Cases") +
  scale_color_discrete(name="") +
  labs(subtitle = "Prediction = Fit exponential function")
```


Today:
```{r echo=FALSE, message=TRUE}
x %>% filter(date == as.Date(Sys.time())) %>% as.data.frame() %>% print()
```


Next three days:
```{r echo=FALSE, message=TRUE}
x %>% filter(date >= as.Date(Sys.time())+1 &
               date <= as.Date(Sys.time())+3 ) %>% as.data.frame() %>% print()
```


# Mortality rate
The following plot shows the mortality rate in all countries with more than 500 cases.
I calculate the mortality rate by dividing the number of COVID19 related deaths by the number of reported cases.

A word of caution: 
Since many reports suggest a high percentage of infected people which are asymptomatic, the number of reported infections are underestimating the number of truely infected.
The high mortality rate in Italy might be explained by Italy's [ways of testing for COVID19](https://medium.com/@andreasbackhausab/coronavirus-why-its-so-deadly-in-italy-c4200a15a7bf) and is probably strongly overestimating the mortality rate.

```{r, dpi=300, fig.width=10, fig.height=7}
dr %>%
  filter(cases > 500) %>%
  group_by(`Country/Region`) %>%
  filter(mean(death_rate) > 0) %>%
  ungroup() %>%
  ggplot(aes(x = date,
             y = death_rate * 100,
             color=`Country/Region`)) +
  geom_line(size=.25) +
  geom_point(size=.1)+
  facet_wrap(vars(`Country/Region`), scales = "free_y") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5),
        legend.position = "none") +
  ylab("Mortality rate [%]")
```

